Density-dependent selection and the limits of relative fitness

### 2018-03-20 manuscript.tex
l

In practice, many population genetic models simply assign a constant relative fitness to each type, and ignore density dependence. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations. A major focus of previous studies in this area is the evolution of traits that affect the ability to keep growing at higher densities \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}. But more broadly, density changes in response to selection in many models of density-regulated growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.

@@

In light of this difficulty, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

The preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}),

%This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.

much of the later reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957} rather than the issues surrounding the constant-$s$ approximation.

For instance, \cite{kimura1969natural} simply assumed that $N$ should remain constant

as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.

In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap. 4]{charlesworth_1994} (e.g. adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group).

  As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

\citep{charlesworth_1971,kimura_1978,leon_1978,nagylaki_1979}

Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). The classic lottery also assumes constant $N$, and generations can overlap. Our first task is to relax the constant-$N$ requirement of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').

>>>>

In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations. Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. Indeed, selection in crowded conditions is intimately connected to equilibrium density in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).

@@

In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type can be justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

The preceding arguments are well-known, and appear in various forms in standard textbooks (e.g. \citealt{crow_1970,begon_1990,nagylaki_1992,otto_2011,ewens_2004}). However, the constant relative fitness idealization of population genetics need not \textit{only} apply when selection is weak (or when selection in genuinely density-independent). Since this idealization is so powerful, it is important to understand the specifics of how and when it fails when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

The models of density-regulated growth mentioned so far are all missing an aspect of selection with major ramifications for the constant-$s$ description of selection: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses. \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types. In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the above models without age structure \citep[Chap. 4]{charlesworth_1994}. For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp. 148]{charlesworth_1994}. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is still largely verbal.

Here we investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess. We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions. Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}. The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g. \citealt[pp. 185]{crow_1970}). However, the classic lottery also assumes constant $N$. Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
